Composite Bayesian inference

### 2018-11-30 notes/composite_inference.tex
Update notes on conditional maxent
Modifs 22/11/18
Modifs 30/11/18

D\`es lors, il est naturel de chercher \`a minimiser le {\em pire regret}: 
$$
R(q) = \max_{y, \theta} L(y, \theta; q)
$$ 
en notant au passage qu'on ne fait pas l'hypoth\`ese que $y$ est distribu\'e selon un certain $p_\theta(y)$. Cette approche \'evite de choisir des a priori sur $y$ et $\theta$. Comme c'est classique dans ce type de probl\`emes minimax, on probabilise $y$ en introduisant le risque moyen:
$$
L(p, \theta; q) 
= E_p [L(y, \theta; q)] 
= \int p(y) \log \frac{p_\theta(y)}{q(y)} dy 
= D(p\|q) - D(p\|p_theta) 
$$

@@

Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas p, la ``vraie'' distribution des donn\'ees, et que $S_\star$ est donc inconnu. Mais Akaike justifie l'approximation: 
$$S_\star \approx \log \ell_\star(y) - k$$ 
o\`u k est la dimension de $\theta$. Tr\`es franchement, j'ai du mal \`a suivre son raisonnement... 
J'ai l'impression qu'il suffit d'appliquer le fameux th\'eor\`eme de Wilks: 
$$2 \log \frac{\ell_\star(y)}{p(y)} \sim \chi_2(k)$$ 
valable asymptotiquement {\em si $p(y)$ appartient \`a la famille} et sous les conditions usuelles d'\'echantillonnage simple. Wilks nous dit alors que la log-vraisemblance profile sur-estime le score du mod\`ele par la moiti\'e d'un $\chi_2$ \`a $k$ degr\'es de libert\'e. En prenant l'esp\'erance de l'expression ci-dessus, les choses deviennent tr\`es claires: 
$$S_\star = E[\log \ell_\star(y)] - \frac{k}{2}$$

@@

Or, si on essaie d'apprendre les poids plut\^ot que de les choisir {\em a priori}, il faudrait alors prendre des poids adaptatifs selon ma th\'eorie... pour le r\'esultat qu'on conna\^it. Conclusion: la vraisemblance composite n'a aucun sens.

C'est bien s\^ur absurde. Si, par exemple, les attributs sont vraiment ind\'ependants, cette m\'ethode fournit une pond\'eration totalement sous-optimale. Car elle suppose implicitement une sorte de "redondance maximale" entre les attributs qui n'est pas la r\'ealit\'e.

@@

Plusieurs remarques s'imposent:
\begin{enumerate}
    \item L'aggr\'egation d'opinions n'est autre qu'un classifieur MaxEnt d'un genre un peu particulier o\`u les contraintes de moments incorporent des mod\`eles g\'en\'eratifs plut\^ot que des quantit\'es arbitraires.
    \item  Une fois r\'ealis\'e le premier apprentissage, le deuxi\`eme apprentissage est d\'etermin\'e par les distributions g\'en\'eratives apprises et par la distribution marginale empirique $h(y)$, cf.~ma note {\tt oct14.pdf}.
    \item On peut voir le MaxEnt conditionnel comme la recherche d'une distribution jointe $p(x,y)$. Il n'y a pas unicit\'e de la solution. Toute distribution de la forme $h(y)p_\star(x|y)$, o\`u $h(y)$ est la marginale d'une distribution $h(x,y)$ compatible avec les contraintes, est optimale.
    \item Le choix d'un {\em a priori} sur $x$ est n\'ecessaire pour d\'efinir les contraintes. Il est naturel de choisir $\pi(x)$ (qui est uniforme dans la pr\'esentation de Berger), ce qui revient \`a poser que la distribution empirique des donn\'ees ajust\'ee pour avoir $h(x)=\pi(x)$ est dans l'espace de recherche.
    \item Toutefois, la m\'ethode n'impose pas que les distributions admissibles aient $\pi(x)$ comme loi marginale. En d'autres termes, tout mod\`ele ayant un fort pouvoir pr\'edictif est susceptible de nous int\'eresser qu'il utilise un {\em a priori} farfelu ou pas. C'est pragmatique \`a d\'efaut d'\^etre bay\'esien.
    \item Cela \'etant, on peut voir la r\`egle d'aggr\'egation comme une g\'en\'eralisation de la loi de Bayes dans laquelle $\pi(x)$ joue le r\^ole d'{\em a priori}. On pourrait dire un {\em a priori} composite.
    \item Une variante du MaxEnt, le principe MIL (Yuan et Clarke, 1999), permet d'imposer la contrainte $\pi(x)=p(x)$. Elle conduit \`a estimer en alternance les distributions marginale et conditionnelle $p(y)$ et $p(y|x)$ via l'algorithme de Blahut-Arimoto. Cette approche est adapt\'ee \`a des contraintes de moments arbitraires mais aurait peu de sens coupl\'ee avec le pr\'e-apprentissage "g\'en\'eratif". Le MIL est une s\'election de mod\`ele pour une inf\'erence authentiquement bay\'esienne et non composite.
\end{enumerate}

@@

@@

\section{Maximum entropie}

>>>>

D\`es lors, il est naturel de chercher \`a minimiser le {\em pire regret}: 
$$
R(q) = \max_{y, \theta} L(y, \theta; q)
$$ 
en notant au passage qu'on ne fait pas l'hypoth\`ese que $y$ est distribu\'e selon un certain $p_\theta(y)$. Cette approche \'evite de choisir des a priori sur $y$ et $\theta$. Comme c'est classique dans ce type de probl\`emes minimax, on probabilise $y$ en introduisant le risque moyen:
$$
L(p, \theta; q) 
= E_p [L(y, \theta; q)] 
= \int p(y) \log \frac{p_\theta(y)}{q(y)} dy 
= D(p\|q) - D(p\|p_\theta) 
$$

@@

Tout semble parfait jusqu'au moment o\`u on r\'ealise qu'on ne ne connait pas $p$, la ``vraie'' distribution des donn\'ees, et que $S_\star$ est donc inconnu. Mais Akaike justifie l'approximation: 
$$S_\star \approx \log \ell_\star(y) - k$$ 
o\`u $k$ est la dimension de $\theta$. Tr\`es franchement, j'ai du mal \`a suivre son raisonnement... 
J'ai l'impression qu'il suffit d'appliquer le fameux th\'eor\`eme de Wilks: 
$$2 \log \frac{\ell_\star(y)}{p(y)} \sim \chi_2(k)$$ 
valable asymptotiquement {\em si $p(y)$ appartient \`a la famille} et sous les conditions usuelles d'\'echantillonnage simple. Wilks nous dit alors que la log-vraisemblance profile sur-estime le score du mod\`ele par la moiti\'e d'un $\chi_2$ \`a $k$ degr\'es de libert\'e. En prenant l'esp\'erance de l'expression ci-dessus, les choses deviennent tr\`es claires: 
$$S_\star = E[\log \ell_\star(y)] - \frac{k}{2}$$

@@

Or, si on essaie d'apprendre les poids plut\^ot que de les choisir {\em a priori}, il faudrait alors prendre des poids adaptatifs selon ma th\'eorie... pour le r\'esultat qu'on conna\^it. C'est bien s\^ur absurde. Si, par exemple, les attributs sont vraiment ind\'ependants, cette m\'ethode fournit une pond\'eration totalement sous-optimale. Car elle suppose implicitement une sorte de "redondance maximale" entre les attributs qui n'est pas la r\'ealit\'e.

@@

Plusieurs remarques s'imposent:
\begin{enumerate}
    \item L'aggr\'egation d'opinions n'est autre qu'un classifieur MaxEnt d'un genre un peu particulier o\`u les contraintes de moments incorporent des mod\`eles g\'en\'eratifs appris au pr\'ealable plut\^ot que des quantit\'es arbitraires.
    \item  Une fois r\'ealis\'e le premier apprentissage, le deuxi\`eme apprentissage est d\'etermin\'e par les distributions g\'en\'eratives apprises et par la distribution marginale empirique $h(y)$, cf.~ma note {\tt oct14.pdf} qui contient par ailleurs des erreurs.
    \item On peut voir le MaxEnt conditionnel comme un MaxEnt relatif standard sur une distribution jointe de marginale connue $h(y)$, donc de la forme $p(x,y)=h(y)p(x|y)$, et par rapport \`a une loi de r\'ef\'erence de la forme $\pi(x,y)=\pi(x)h(y)$. Il est donc capital d'avoir une estim\'ee de la marginale $h(y)$.
    \item Le choix d'un {\em a priori} sur $x$ est n\'ecessaire pour d\'efinir les contraintes. Il est naturel de choisir $\pi(x)$ (qui est uniforme dans la pr\'esentation de Berger), ce qui revient \`a poser que la distribution empirique des donn\'ees ajust\'ee pour avoir $h(x)=\pi(x)$ est dans l'espace de recherche.
    \item L'espace de recherche est d\'efini par des contraintes portant sur les log-vraisemblances moyennes  qui sont {\em moins fortes} que supposer connues les distributions des attributs. Elles expriment l'id\'ee tr\`es MDL-esque que ce qui est important avec un mod\`ele, ce n'est pas qu'il soit exact, mais le taux de compression qu'il permet d'atteindre. 
    \item Doit-on utiliser des contraintes d'\'egalit\'e ou d'in\'egalit\'e? Les premi\`eres sont plus naturelles mais permettent des poids n\'egatifs... ce qui est malvenu pour des ``opinions''. 
    \item Toutefois, la m\'ethode n'impose pas que les distributions admissibles aient $\pi(x)$ comme loi marginale. En d'autres termes, tout mod\`ele ayant un fort pouvoir pr\'edictif est susceptible de nous int\'eresser qu'il utilise un {\em a priori} farfelu ou pas. C'est pragmatique \`a d\'efaut d'\^etre bay\'esien.
    \item Cela \'etant, on peut voir la r\`egle d'aggr\'egation comme une g\'en\'eralisation de la loi de Bayes dans laquelle $\pi(x)$ joue le r\^ole d'{\em a priori}. On pourrait dire un {\em a priori} composite.
    \item Une g\'en\'eralisation du MaxEnt conditionnel, le MIL (Yuan et Clarke, 1999) permet de rel\^acher la contrainte que $p(y)$ est connue et (\'eventuellement) d'imposer la contrainte $\pi(x)=p(x)$. Elle conduit \`a estimer en alternance les distributions marginale et conditionnelle $p(y)$ et $p(y|x)$ via l'algorithme de Blahut-Arimoto. \c Ca semble indiqu\'e si on n'a pas confiance dans la distribution empirique $h(y)$ mais, dans ce cas, pourquoi avoir confiance dans les moments? 
\end{enumerate}

@@

\section{Compression}

Let $X\sim p_\star(x)$ and consider a model $q(x)$. Assume we know the compression rate of $q$, but don't know $p_\star$. Does it make sense to consider admissible any distribution $p(x)$ for which the compression rate is at least as large, hence satisfying the following inequality:
$$
L(p,q) \geq L(p_\star,q),
$$
with:
$$
L(p,q) = \int p(x)\log q(x) dx 
$$

This is weird because if we somehow know $L(p_\star,q)$, we know for sure that it can't be larger than what it is, right? So why not using an equality constraint? Unless what we think we know happens to be a lower bound.

$$
\max_q\min_{p\in P} L(p,q) 
=
\min_{p\in P} \max_q L(p,q)
= \min_{p\in P} L(p,p)
= -\max_{p\in P} H(p)
$$

@@

\section{Maximum entropie conditionnelle}

$$
D(p) = \int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
$$

Lagrangian:
$$
{\cal L}(p,\lambda) = D(p) -\lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - F)
$$

Optimal $p$ at fixed $\lambda$:
$$
\frac{\partial{\cal L}}{\partial p}
=
h(y) \log \frac{p(x|y)}{\pi(x)} - h(y) \lambda^\top f(x,y)
= h(y) c(y)
\quad
\Rightarrow
\quad
p_{\lambda}(x|y) = \frac{\pi(x)}{z(\lambda,y)} e^{\lambda^\top f(x,y)}
$$
with:
$$
z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
$$

Dual:
$$
\psi(\lambda) = \lambda^\top F - \int h(y) \log z(\lambda, y) dy 
$$

$$
\nabla z
=
\int \pi(x) f(x,y)  e^{\lambda^\top f(x,y)} dx
$$

$$
\nabla \nabla^\top z 
=
\int \pi(x) f(x,y) f(x,y)^\top e^{\lambda^\top f(x,y)} dx
$$

$$
\nabla\psi
= F - \int h(y) \frac{\nabla z}{z} dy
$$

$$
\nabla\nabla^\top\psi
= - \int h(y) \left(
\frac{\nabla \nabla^\top z}{z} 
-
\frac{\nabla z \nabla z^\top}{z^2}
\right)
dy
$$

Note the equivalent expression of the dual function as:
$$
\psi(\lambda)
=
\int h(x,y) \log \frac{p_\lambda(x|y)}{\pi(x)} dx dy
= D(h\|\pi) - D(h\|p_\lambda)
$$

\section{Maxent via the generalized KL}

It's a bit different as the unit integral constraint is made implicit.

Assume $\int\pi(x)dx=1$ and $\int h(y) dy = 1$. 
$$
D(p) 
= \int h(y)p(x|y) \log \frac{p(x|y)}{\pi(x)} dxdy
- \int h(y)p(x|y)dx dy
+ 1
$$

Goal is to minimize $D(p)$ s.t. mean-value constraints on $p$.

Lagrangian:
$$
{\cal L}(p,\lambda) = D(p) -\lambda^\top (\int h(y)p(x|y)f(x,y)dxdy - F)
$$

Optimal $p$ at fixed $\lambda$:
$$
\frac{\partial{\cal L}}{\partial p}
=
h(y) \log \frac{p(x|y)}{\pi(x)} - h(y) \lambda^\top f(x,y)
= 0
\quad
\Rightarrow
\quad
p_{\lambda}(x|y) = \pi(x) e^{\lambda^\top f(x,y)}
$$

Dual:
$$
\psi(\lambda) = \lambda^\top F - \int h(y)z(\lambda, y) dy + 1
$$
with:
$$
z(\lambda,y) = \int \pi(x) e^{\lambda^\top f(x,y)} dx
$$

$$
\nabla\psi(\lambda)
= F - \int \pi(x) h(y) f(x,y) e^{\lambda^\top f(x,y)} dx dy
$$

$$
\nabla\nabla^\top\psi(\lambda)
= - \int \pi(x) h(y) f(x,y)f(x,y)^\top e^{\lambda^\top f(x,y)} dx dy
$$

### 2018-11-30 papers/composite_inference/draft.tex
Modifs 30/11/18

Classical frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable. Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate the field of artificial intelligence and typically require heavy supervised learning -- these include classical techniques such as maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}, as well as most of deep learning \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06,Fischer-14}).

@@

\section{Maximum entropy derivation of composite likelihood}
\label{sec:maxent}

The coordinator discards the priors used by the agents (if non-uniform).

@@

BTW, we can work with inequality constraints to force $\lambda\geq 0$:
$$
E[\log p(z_i|x)] \geq c_i \equiv \int p(z_i|x) \log p(z_i|x) dx
$$

@@

Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion ofR feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?

@@

The dual function amounts to the famous cross-entropy.

@@

>>>>

Classical frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable. Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate the field of artificial intelligence and typically require massive supervised learning -- these include classical techniques (maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.

@@

\section{Maxent composite likelihood}
\label{sec:maxent}

We can derive composite likelihood from the conditional maximum entropy principle \cite{BergerA-96}. It comes with a method of tuning weights, hence a particular composite likelihood function that we call maxent composite likelihood (MCL).

@@

We work with inequality constraints to force $\lambda\geq 0$:
$$
E[\log p(z_i|x)] \geq c_i \equiv \int \pi(x)p(z_i|x) \log p(z_i|x) dx dy
$$

@@

Why not using the feature-based posteriors or Bayes factors rather than the likelihoods? It's a possibility (perhaps a connection here with Gr\"unwald's luckiness). A justification for not doing it is that the coordinator should discard the priors used by the agents. The constraints are not purely evidence-based since the moments depend on the coordinator's prior but it is arguable a mess of priors wouldn't be good.

Intuition behind weights: Lagrange multipliers to enforce a constraint... Notion of feature relevance. But does $\lambda_1>\lambda_2$ imply that $z_1$ is more relevant than $z_2$?

The derivation assumed the marginal distribution of the data $h(y)$ to be known. In practice, it is estimated by the empirical distribution just like the moments are estimated.

@@

The dual function boils down to the famous cross-entropy.

Single feature case ($y=z$): the maxent solution has the form $p(x|y)\propto\pi(x)p(y|x)^\lambda$. We expect to have $\lambda=1$ but it's not necessarily the case. It is if $h(y)=\int\pi(x)p(y|x)dx$ because then $h(y)p(x|y)=\pi(x)p(y|x)$ so the constraint is verified (and active). This also tells us that the data marginal should be consistent with the prior on labels for this much expected consistency property to hold. So, either weight datapoints so as to match a desired prior (e.g., uniform) or take the empirical distribution of labels as the ``prior''.

Which brings another question: in this case, the (unique) weight is insensitive to the prior -- is it true in general? The answer is no. The weights generally depend on the prior. Hence the maxent composite likelihood is prior-dependent. This is an important conceptual difference with the classical notion of likelihood.

\section{Composite EM algorithm for unsupervised learning}
\label{sec:gem}

In a nutshell: alternate unsupervised learning of partial model parameters ($M$-step) and re-estimation of label posteriors via maxent ($E$-step).

$M$-step: given a posterior estimate $q(x|y)$, do just as if the features were independent and solve: 
$$
\max_\theta
\int \sum_i h(y)q(x|y) \log p_\theta(z_i,x) dx dy
$$

Note we can estimate cross-feature parameters such as class proportions.

Surrogate $E$-step: given generative parameters, recompute the moment constraints and update $q(x|y)$ according to the conditional maxent principle:
$$
\min_q
\int h(y)q(x|y) \log \frac{q(x|y)}{\pi(x)} dx dy
$$
s.t.
$$
\int h(y)q(x|y)\log p_\theta(z_i|x) dx dy \geq \int \pi(x)p_\theta(z_i|x)\log p_\theta(z_i|x) dx dy
$$

However, there is an important issue: how is $h(y)$ defined in both the $M$- and $E$-steps? In supervised context, $h(x,y)$ can be defined beforehand as the prior-corrected empirical distribution:
$$
h(x,y) = \pi(x) \frac{h_0(x,y)}{h_0(x)},
$$
where $h_0(x,y)$ is the raw empirical distribution. In unsupervised context, the best we can do is to replace $h_0(x,y)$ with $h_0(y)q(x|y)$, implying that the empirical distribution of $y$ is reweighted according to:
$$
h(y) = \sum_i w_i\delta(y-y_i),
\qquad
w_i = \int \pi(x)\frac{q(x|y_i)}{\sum_j q(x|y_j)} dx,
$$
which needs to be done each time $q(x|y)$ is updated, that is, after each $E$-step.

The other option is to keep $h(y)$ constant throughout the iterations but adapt $\pi(x)$ as the marginal distribution of labels learned in the $M$-step. In this case, it's no longer a ``prior''.

Are we minimizing a unique functional as in the variational EM algorithm? Possibly not. We may see this generalization as a two-player game. One player tries to predict several data features independently. The other player tries to predict labels. Together, they find patterns (clusters) in unlabeled data. Does the more general version converge? Does the game have a value? And so on.

In the case of a single feature, is this the well-known EM? The $M$-step is clearly the same, what about the $E$-step? It is the same if $h(y)p_\theta(x|y)=p_\theta(y|x)\pi(x)$. However, this condition won't hold if $h(y)$ is an empirical distribution... In this special case, the ``E player'' could use the data distribution model produced by the ``M player'' instead of the empirical one but, more generally, there is no data distribution model so he has to stick to the empirical one.

The maxent is a surrogate for the $E$-step. It reflects the lack of a full model $p_\theta(x,y)$.

@@

Deep learning is discriminative...
